---
title: "Lab 4 Distribution Free tests"
output: html_document
date: "2024-09-13"
---

## Resourcecs for todays lab




## Distribution Free Tests

Last week we covered t-tests. T tests examine differences in means based upon assumtions of their distribution. 
Today we will be covering tests that do make assumptions on distribution. These tests are also called non-parametric tests.
These tests are often used with smaller sample sizes and rely on medians or ranks of data rather than means. 

We will be covering 3 tests today:
1. Wilcoxon signed rank - single group test (i.e., like the single sample t-test)
2. Wilcoxon matched pairs - paired test (i.e., like the paired sample t-test)
3. Mann-Whitney U - two sample (i.e., like the two sample t test)

### Set up
Load packages
```{r}
if(!require(tidyverse)) {install.packages("tidyverse")}
library(tidyverse) # for data stuff
if(!require(haven)) {install.packages("haven")}
library(haven) # for importing spss files
if(!require(report)) {install.packages("report")}
library(report) # for fancy t test reporting
```

### Import our data
```{r}
# select the dataset from homework 1
df <- read_sav(file.choose())
```

### Prep our data
```{r}
df$male = as.factor(df$male)
df$group = as.factor(df$group)
df
```



Wilcoxon matched pairs 
```{r}
a = wilcox.test(y, y2, paired = TRUE)

```






