---
title: "Lab 4 Distribution Free tests"
output: html_document
date: "2024-09-13"
editor_options: 
  markdown: 
    wrap: 72
---

# Lab 4 Distribution Free Tests

## Information about lab

Last week we covered t-tests. T tests examine differences in means based
upon assumtions of their distribution. Today we will be covering tests
that do make assumptions on distribution. These tests are also called
non-parametric tests. These tests are often used with smaller sample
sizes and rely on medians or ranks of data rather than means.

We will be covering 4 tests today:

Alternatives to single sample t tests

1.  Binomial Sign Test
2.  Chi square goodness of fit

Alternative to two sample t tests

3.  Moods median
4.  Wilcoxon matched pairs - paired test (i.e., like the paired sample
    t-test)

Tests we are not going over but I'm providing references for:

-   Wilcoxon signed rank - single group test (i.e., like the single
    sample t-test, but with a median).

-   Wilcoxon signed rank - paired test (i.e., like the paired sample
    t-test, but with a median).

## References

Man whitney U test (i.e., wilcoxon tests) -
<https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/> Making
figures with ggpubr -

## Set up

### Packages

install packages

```{r, results= 'hide', message=FALSE, warning=FALSE}
if(!require(tidyverse)) {install.packages("tidyverse")}
if(!require(haven)) {install.packages("haven")}
if(!require(rstatix)) {install.packages("rstatix")}
if(!require(coin)) {install.packages("coin")}

# install.packages('effectsize')
```

load packages

```{r, results= 'hide', message=FALSE, warning=FALSE}
library(tidyverse) # data stuff
library(dplyr) #data stuff
library(haven) # for importing spss files
library(rstatix) # for wilcoxan stuff
library(report)
library(ggpubr) # for plots
library(coin)
library(gt)
# library(effectsize)
```

### Import our data

```{r}
path <- file.choose()
df <- read_sav(path)
```

### Prep our data

```{r}
df$male = as.factor(df$male)
df$male <- recode(df$male, "1" = "Female", "2" = "Male")
df$group = as.factor(df$group)
df2 <- filter(df,id < 21)
#df 2 is a sample of only 20 people, df is the sample with all of the people in it.
df2
```

## 1. Binomial Test - Nonparametic tests

The binomial test assesses if there are equal number oof people above
and below the mean. We are interested in seeing if equal number of
people are above and below a depression score of 26, as this is the
"mean" of the general population. We are using df2 as this is the
dataframe with a smaller sample.

1.  See how many people are in each group

```{r}
# create a variable of scores above or below 26 and make a table
df2$success<- ifelse(df2$depression > 25,"above","below") 
#table
table(df2$success)
#length
length <- length(df2$success)
names(length) = "n"
length
```

You can see out of 2 people have a score of 25 or below with a total
number of 20 people in the sample. We would suspect that there is a
difference here.

2.  Running the Binomial test

```{r}
# template for this is binom.test(number of yeses, total amount)
binom.test(x = 18,n =20, p =.5)

```

Just like SPSS, Its significant! We reject the null hypothesis that we
the groups are split 50/50.

3.  Writing it up - report doesn't work for this one, sorry! Based on a
    one-sided binomial test the observed values (N= 20,K= 18) give
    significant support to the assumption that more than 50% of the
    sample is depressed, p= \<.001.

## 2. Chi Square - single variable vs population

### Two group Chi square - Single sample vs population

Let’s also say that the gender makeup of college students is 57% female
Does our sample have an equal gender breakdown as the larger population?

1.  See how many people are in each group

```{r}
table(df$male)
plot(df$male)
```

2.  Run and report the test

```{r}
#take the output from the frequency table and assign it to a variable male
male<-c(335, 106)
#run the chi square with the 57% female and 43% male expectation
chisq.test(male, p = c(.57, .43))
#use report to tell us what this means!
report(chisq.test(male, p = c(.57, .43)))
```

### Three Group Chi Square - Single sample vs population

We’re gonna use the variable ‘group’ Let’s say that group should be a
variable where the distribution is 20/20/60.

1.  Run the test for the 3 groups to see if there is a significant
    difference.

```{r}
#get the frequencies
table(df$group)
plot(df$group)
#assign to group variable
group<-c(28, 62, 351)
#run the test
chisq.test(group, p = c(.2, .2, .6))
#report the test!
report(chisq.test(group, p = c(.2, .2, .6)))
```

Now we know that there is a significant difference. We have to do
separate tests to examine differences between each group.

The proportions between the groups: Group 1 v. Group 2 (proportion
50/50) Group 1 v. Group 3 (proportion 1:3, 25/75) Group 2 v. Group 3
(proportion 1:3, 25/75)

1.  Start with frequencies

```{r}
#We don't need to filter our variables, like SPSS we just make new Group variables
table(df$group)
```

2.  Assign values to group variables,

```{r}
groups_1_2 <- c(28,62)
groups_1_3 <- c(28,351)
groups_2_3 <- c(62,531)
```

3.  Run the tests separately with proportions above (what we expect)
    Between Groups 1 and 2

```{r}
chisq.test(groups_1_2, p = c(.5, .5))
report(chisq.test(groups_1_2, p = c(.5, .5)))
```

4.  Between groups 1 and 3

```{r}
chisq.test(groups_1_3, p = c(.25, .75))
report(chisq.test(groups_1_3, p = c(.25, .75)))
```

5.  Between groups 2 and 3

```{r}
chisq.test(groups_2_3, p = c(.25, .75))
report(chisq.test(groups_2_3, p = c(.25, .75)))
```

We can see that there are significant differences in the proportions of
students in groups compared to what we would expect in the population!

## 3. Moods Median Test - two groups

## 4. Man Whitney U (aka Wilcox test) - two groups

This test serves as a distribution free analog of the two sample t test.

1.  Examine differences between the two groups

```{r}
# "%>% these are called pipe operators it basically lets us pass information to the next line. 
df2 %>%
  group_by(male) %>%
  get_summary_stats(depression, type = "median_iqr")
```

2.  Run the test

```{r}
# run stats
wilcox.test(df2$depression ~ df2$male)
# effect size
df2 %>% wilcox_effsize(depression ~ male) 
# z test
z = as.numeric(statistic(wilcox_test(df2$depression ~ df2$male), type="standardized"))
names(z) = "z"
z

#save stats
results2 <- df2 %>% 
  rstatix::wilcox_test(depression ~ male) %>%
  add_significance() %>%
  add_xy_position(x = "male")
results2_es <- df2 %>% wilcox_effsize(depression ~ male) # runs the effect size
```

3.  Report it!

We could report the result as follow:

A Wilcoxon signed-rank test was computed to assess whether depression
differs by sex within a student sample. The median depression in female
group was 44 (IQR = 18.50), whereas the median in male group was 39.9
(IQR = 15.25). The Wilcoxon test showed that the difference was not
significant (z = -0.36, p = 0.73, effect size r = 0.06).

4.  Visualization

```{r}
#simple plot
ggboxplot(df2,
  y =  "depression", # outcome variable
  x =  "male", # predictor variable
  ylab = "Depression",  #label your y axis
  xlab = "Gender", # label your x axis,
  color = "male"
)

#fancy plot
ggboxplot(df2,
  y =  "depression", # outcome variable
  x =  "male", # predictor variable
  ylab = "Depression",  #label your y axis
  xlab = "Gender", # label your x axis,
  # color = "male",
  #leave the rest below the same
  width = .5, 
  add = c("median", "jitter"), 
  add.params = list(color= "male",alpha = .5)) + # the plus sign lets us add more things to our plot when using a ggplot
  stat_pvalue_manual(results2, tip.length = 0) +
  labs(
    caption = # we are making a subtitle
    paste( #manually put in the info that you want!
      "Wilcoxon test,", "z =", round(z,2), "p =", round(results2$p,4),
      ", r =", round(results2_es$effsize,2)))
```

## Paired sample wilcox test

1.  Summary statistics

lets look at the difference between pre and post scores of depression

```{r}
df %>%
  get_summary_stats(c(depression, depression2), type = "median_iqr")
```

1.1 we have to run this for some of the stats to work. I am happy to
teach you if you want to learn more about it

```{r}
df_long <- df %>% 
  pivot_longer(
    cols = depression:depression2,
    names_to = "Timepoint",
    values_to = "Depression") %>%
  mutate(
    Timepoint = as.factor(Timepoint),
    Timepoint = recode(
      Timepoint,
      "depression" = "pre",
      "depression2" = "post"))
```

2.  Run the tests

```{r}
# test
wilcox.test(df$depression, df$depression2, paired = TRUE)
# effect size
df_long %>% wilcox_effsize(Depression ~ Timepoint, paired = TRUE)

# save test

results3 <- df_long %>% 
  rstatix::wilcox_test(Depression ~ Timepoint, paired = TRUE) %>%
  add_significance() %>%
  add_xy_position(x = "Timepoint")
results3_es <- df_long %>% wilcox_effsize(Depression ~ Timepoint, paired = TRUE)
```

3.  Report the findings

A paired sample Wilcoxon signed rank test was conducted to example
differences pre and post depression treatmen. The median depression of
the participants before treatment is significantly higher (M = 40) from
the median depression score after treatment (M = 27) with a p-value \<
0.001 and effect size of r = 0.87.

4.  Visulization

```{r}
ggpaired(df_long, x = "Timepoint", y = "Depression", 
         order = c("pre", "post"),
         ylab = "Depression", xlab = "Timepoint",
        # extra stuff
         point.size = 1.2,
         line.color = "grey",
         color = "blue") +
         stat_pvalue_manual(results3, tip.length = 0) +
         labs(
          caption = # we are making a subtitle
              paste( #manually put in the info that you want!
      "Wilcoxon test,", "V =", results3$statistic, "p =", round(results3$p,4),
      ", r =", round(results3_es$effsize,2)))
        
```
