---
title: "Lab 4 Distribution Free tests"
output: html_document
date: "2024-09-13"
editor_options: 
  markdown: 
    wrap: 72
---

## Resourcecs for todays lab

Running wilcoxon tests

\- <https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/> Making
figures with ggpubr

## Distribution Free Tests

Last week we covered t-tests. T tests examine differences in means based
upon assumtions of their distribution. Today we will be covering tests
that do make assumptions on distribution. These tests are also called
non-parametric tests. These tests are often used with smaller sample
sizes and rely on medians or ranks of data rather than means.

We will be covering 3 tests today:

1\. Wilcoxon signed rank - single group test (i.e., like the single
sample t-test)

2\. Wilcoxon matched pairs - paired test (i.e., like the paired sample
t-test)

3\. Mann-Whitney U - two sample (i.e., like the two sample t test)

### Set up

Load packages

```{r, results= 'hide', message=FALSE, warning=FALSE}
if(!require(tidyverse)) {install.packages("tidyverse")}
if(!require(haven)) {install.packages("haven")}
if(!require(rstatix)) {install.packages("rstatix")}
if(!require(coin)) {install.packages("coin")}

# install.packages('effectsize')
```

load packages

```{r, results= 'hide', message=FALSE, warning=FALSE}
library(tidyverse) # data stuff
library(dplyr) #data stuff
library(haven) # for importing spss files
library(rstatix) # for wilcoxan stuff
library(ggpubr) # for plots
library(coin)
library(gt)
# library(effectsize)
```

### Import our data

```{r}
# select the dataset from homework 1

path <- file.choose()
df <- read_sav(path)
```

### Prep our data

```{r}
df$male = as.factor(df$male)
df$male <- recode(df$male, "1" = "Female", "2" = "Male")
df$group = as.factor(df$group)
df <- filter(df,id < 30)
df
```

## Single Sample Wilcoxon

### 1. Descriptive for our question

```{r}
# the %>% symbol is called a pipe operator and its "pipes" or moves data from one line to another.You'll see why we use this as we keep going. 
df %>% 
  get_summary_stats(depression, type = "median_iqr")
```

### 2. Run the test

```{r}
library(rstatix)
#runs the tests 
wilcox.test(df$depression ~ 1, mu = 25)
#runs the effect size
df %>% wilcox_effsize(depression ~ 1, mu = 25) # runs the effect size

#median
median(df$depression)

#save these for the plot below
w1_es <- df %>% wilcox_effsize(depression ~ 1, mu = 25)
single_wilcox <- df %>% rstatix::wilcox_test(depression ~ 1, mu = 25)
```

### 3. Report it!

We could report the result as follow:

A Wilcoxon signed-rank test was computed to assess whether the sample
median depression was different to the population normal median
depression (25). The measured student median depression (40) was
statistically significantly lower than the population median depression
of 25 (p \< 0.001, effect size r = 0.88).

### 4. Visualization

```{r}
ggboxplot(
  df$depression, # variable 
  ylab = "Depression",  #label your y axis
  xlab = "College Students", # label your x axis
  #leave the rest below the same
  width = .5, 
  add = c("median", "jitter"), 
  add.params = list(color= "blue",alpha = .5))+ # the plus sign lets us add more things to our plot when using a ggplot
  labs(
    caption = # we are making a subtitle
    paste( #manually put in the info that you want!
      "Wilcoxon test,", "p =", round(single_wilcox$p,4),
      ", r =", round(w1_es$effsize,2)))
```

## Wilcoxon rank sum test - two groups

### 1. Examine differences between the two groups

```{r}
# these are called pipe operators it basically lets us pass information to the next line. 
df %>%
  group_by(male) %>%
  get_summary_stats(depression, type = "median_iqr")
```

### 2. Run the test

```{r}
# run stats
wilcox.test(df$depression ~ df$male)
# effect size
df %>% wilcox_effsize(depression ~ male) 
# z test
z = as.numeric(statistic(wilcox_test(df$depression ~ df$male), type="standardized"))
names(z) = "z"
z

#save stats
results2 <- df %>% 
  rstatix::wilcox_test(depression ~ male) %>%
  add_significance() %>%
  add_xy_position(x = "male")
results2_es <- df %>% wilcox_effsize(depression ~ male) # runs the effect size
```

### 3. Report it!

We could report the result as follow:

A Wilcoxon signed-rank test was computed to assess whether depression
differs by sex within a student sample. The median depression in female
group was 44 (IQR = 18.50), whereas the median in male group was 39.9
(IQR = 15.25). The Wilcoxon test showed that the difference was not
significant (z = -0.36, p = 0.73, effect size r = 0.06).

### 4. Visualization

```{r}
ggboxplot(df,
  y =  "depression", # outcome variable
  x =  "male", # predictor variable
  ylab = "Depression",  #label your y axis
  xlab = "Gender", # label your x axis,
  color = "male"
)

ggboxplot(df,
  y =  "depression", # outcome variable
  x =  "male", # predictor variable
  ylab = "Depression",  #label your y axis
  xlab = "Gender", # label your x axis,
  color = "male",
  #leave the rest below the same
  width = .5, 
  add = c("median", "jitter"), 
  add.params = list(color= "male",alpha = .5)) + # the plus sign lets us add more things to our plot when using a ggplot
  stat_pvalue_manual(results2, tip.length = 0) +
  labs(
    caption = # we are making a subtitle
    paste( #manually put in the info that you want!
      "Wilcoxon test,", "z =", round(z,2), "p =", round(results2$p,4),
      ", r =", round(results2_es$effsize,2)))
```

## Paired sample wilcox test

### 1. Summary statistics

lets look at the difference between pre and post scores of depression

```{r}
df %>%
  get_summary_stats(c(depression, depression2), type = "median_iqr")
```

### 1.1 we have to run this for some of the stats to work. I am happy to teach you if you want to learn more about it

```{r}
df_long <- df %>% 
  pivot_longer(
    cols = depression:depression2,
    names_to = "Timepoint",
    values_to = "Depression") %>%
  mutate(
    Timepoint = as.factor(Timepoint),
    Timepoint = recode(
      Timepoint,
      "depression" = "pre",
      "depression2" = "post"))
```

### 2. Run the tests

```{r}
# test
wilcox.test(df$depression, df$depression2, paired = TRUE)
# effect size
df_long %>% wilcox_effsize(Depression ~ Timepoint, paired = TRUE)

# save test

results3 <- df_long %>% 
  rstatix::wilcox_test(Depression ~ Timepoint, paired = TRUE) %>%
  add_significance() %>%
  add_xy_position(x = "Timepoint")
results3_es <- df_long %>% wilcox_effsize(Depression ~ Timepoint, paired = TRUE)
```

### 3. Report the findings

A paired sample Wilcoxon signed rank test was conducted to example
differences pre and post depression treatmen. The median depression of
the participants before treatment is significantly higher (M = 40) from
the median depression score after treatment (M = 27) with a p-value <
0.001 and effect size of r = 0.87.

### 4. Visulization

```{r}
ggpaired(df_long, x = "Timepoint", y = "Depression", 
         order = c("pre", "post"),
         ylab = "Depression", xlab = "Timepoint",
        # extra stuff
         point.size = 1.2,
         line.color = "grey",
         color = "blue") +
         stat_pvalue_manual(results3, tip.length = 0) +
         labs(
          caption = # we are making a subtitle
              paste( #manually put in the info that you want!
      "Wilcoxon test,", "V =", results3$statistic, "p =", round(results3$p,4),
      ", r =", round(results3_es$effsize,2)))
        
```
