---
title: "Lab 2 t test"
output: html_document
date: "2024-08-29"
---

# References for t-tests and this weeks lab
These are the websites I used to compile information for this week:
You can use this website a reference for t tests:  
- https://www.geeksforgeeks.org/t-test-approach-in-r-programming/
visualizing statistics (i.e., ggstatplot):
- https://indrajeetpatil.github.io/ggstatsplot/
ruporting stats:
- https://easystats.github.io/report/
A reference of most stats and their respective functions:
- https://indrajeetpatil.github.io/statsExpressions/articles/stats_details.html

# T tests
This Markdown will show you code to analyze and report t tests. t tests are used
to compare difereneces between the means. There are 3 t tests we will cover in this class:

1. One sample t tests - comparing the mean a sample to the population mean.
2. Independent t test (two sample t test) - comparing two different groups means.
3. Paired sample t test - comparing group means of people within the same sample.


Today We will be focusing on one sample t test, binomial tests (non-perametic t tests), and chi square tests. 

## Set up
Lets install and load our packages. The fancy code below checks if you have 
installed a package, and if you haven't, it installs it for you.
```{r}
if(!require(tidyverse)) {install.packages("tidyverse")}
library(tidyverse) # for data stuff
if(!require(haven)) {install.packages("haven")}
library(haven) # for importing spss files
if(!require(ggplot2)) {install.packages("ggplot2")}
library(ggplot2) # for fancy t test reporting
if(!require(ggstatsplot)) {install.packages("ggstatsplot")}
library(ggstatsplot) # for fancy t test reporting
if(!require(report)) {install.packages("report")}
library(ggstatsplot) # for fancy t test reporting
```

## Import our data
Lets choose HW1 dataset
```{r}
df <- read_sav(file.choose())
```

# prep our data
```{r}
df$male = as.factor(df$male)
df$group = as.factor(df$group)
df
```

## Single Sample t tests
Using Base R t tests: Lets see if our sample is more depressed than the population:
Again, single sample is comparing the mean of our sample to the population statistic (i.e., mu)
```{r}
# mu is the test statistic mean we are comparing our sample to 
t.test(df$depression, mu = 25)
```

Using the {report} package will tell you if your t-test is significant!
```{r}
library(report)
report(t.test(df$depression, mu = 25))
```

The {ggstatplot} will run the t test for you, show you a histogram of 
your distribution, and show you the mean of your sample, and normal distribution.
```{r}
library('ggstatsplot')
gghistostats(
  data       = df, # Put your data here
  x          = depression,  # put your x variable that you are running the test here
  title      = "Scores of Depression in College Age Students", # change the title
  test.value = 25, # This is the test statistic
  #Don't worry about the stuff below!
  binwidth   = 1,
  effsize.type = "d", # Effect size (i.e., how significant it is)
  normal.curve = TRUE, # adding a normal curve to it
  normal.curve.args = list(color = "darkred",linewidth = 1),
  bf.message = FALSE,
)
```

## Binomial Test - Nonparametic tests

First lets filter out people like we did in spss. And figure out how many people are below 26.
```{r}
#subset
df_small<- subset(df, id < 21)
df_small

#call the variable to see how many people are below 25
hist(df_small$depression)

# OR create a variable and make a table
df_small$success<- ifelse(df_small$depression > 25,"yes","no") 
table(df_small$success)

# this function tells you how many people you have in total
length(df_small$success)
```
You can see out of 2 people have a score of 25 or below.

Running the Binomial test
```{r}
# template for this is binom.test(number of yeses, total amount)
binom.test(18,20)
```
Just like SPSS, Its significant! We reject the null hypothesis that we the groups are split 50/50.

Writing it up: 
Based on a one-sided binomial test the observed values (N = 20, K = 18) give significant support to the assumption that more than 50% of the sample is depressed, p = <.001. 

## Chi Square - single variable

### Two group Chi square
Let’s also say that the gender makeup of college students is 57% female
Does our sample have an equal gender breakdown as the larger population?
```{r}
table(df$male)
```

Plug those numbers in and run the test!
```{r}
#take the output from the frequency table and assign it to a variable male
male<-c(335, 106)
#run the chi square with the 57% female and 43% male expectation
chisq.test(male, p = c(.57, .43))
#use report to tell us what this means!
report(chisq.test(male, p = c(.57, .43)))
```

### three group Chi Square 

We’re gonna use the variable ‘group’
Let’s say that group should be a variable where the distribution is 20/20/60
```{r}
#get the frequencies
table(df$group)
```


```{r}
#assign to group variable
group<-c(28, 62, 351)
#run the test
chisq.test(group, p = c(.2, .2, .6))
#report the test!
report(chisq.test(group, p = c(.2, .2, .6)))
```

We know that there is a significant difference. Now, we just need to find what groups are different from each other!

Here are the proportions we would expect between each of the groups:
Group 1 v. Group 2 (proportion 50/50)
Group 1 v. Group 3 (proportion 1:3, 25/75)
Group 2 v. Group 3 (proportion 1:3, 25/75)
```{r}
#We don't need to filter our variables, just make new Group variables

#1. Start with frequencies 
table(df$group)
```


```{r}
#2. Assign values to group variables, 
groups_1_2 <- c(28,62)
groups_1_3 <- c(28,351)
groups_2_3 <- c(62,531)
```

3. run the tests seperaately with proportions above (what we expect)
Between Groups 1 and 2
```{r}

    chisq.test(groups_1_2, p = c(.5, .5))
    report(chisq.test(groups_1_2, p = c(.5, .5)))
```

Between groups 1 and 3
```{r}
    chisq.test(groups_1_3, p = c(.25, .75))
    report(chisq.test(groups_1_3, p = c(.25, .75)))
```

between groups 2 and 3 
```{r}
    chisq.test(groups_2_3, p = c(.25, .75))
    report(chisq.test(groups_2_3, p = c(.25, .75)))
```

WE can see that there are significant differences in the proportions of students
in groups compared to what we would expect!


