---
title: "Lab 3 t-test"
output: html_document
date: "2024-09-09"
---

# References for t-tests and this weeks lab
These are the websites I used to compile information for this week:
You can use this website a reference for t tests:  
- https://www.geeksforgeeks.org/t-test-approach-in-r-programming/
visualizing statistics (i.e., ggstatplot):
- https://indrajeetpatil.github.io/ggstatsplot/
reporting stats:
- https://easystats.github.io/report/

# T tests
This Markdown will show you code to analyze and report t tests. t tests are used
to compare difereneces between the means. There are 3 t tests we will cover in this class:

1. One sample t tests - comparing the mean a sample to the population mean.
2. Independent t test (two sample t test) - comparing two different groups means.
3. Paired sample t test - comparing group means of people within the same sample.

## Set up
Lets install and load our packages. The fancy code below checks if you have 
installed a package, and if you haven't, it installs it for you.
```{r}
if(!require(tidyverse)) {install.packages("tidyverse")}
library(tidyverse) # for data stuff
if(!require(haven)) {install.packages("haven")}
library(haven) # for importing spss files
if(!require(ggplot2)) {install.packages("ggplot2")}
library(ggplot2) # for fancy t test reporting
if(!require(ggstatsplot)) {install.packages("ggstatsplot")}
library(ggstatsplot) # for fancy t test reporting
if(!require(report)) {install.packages("report")}
library(report) # for fancy t test reporting
```

## Import our data
Lets choose HW1 dataset
```{r}
path <- file.choose()
path
df <- read_sav(path)
```

# prep our data
```{r}
df$male = as.factor(df$male)
df$male <- recode(df$male, "1" = "Female", "2" = "Male")
df$group = as.factor(df$group)
df
```


## Single Sample t tests
Using Base R t tests: Lets see if our sample is more depressed than the population:
Again, single sample is comparing the mean of our sample to the population statistic (i.e., mu)
```{r}
# mu is the test statistic mean we are comparing our sample to 
Single_t<- t.test(df$depression, mu = 25) # the <- assigns the output of that function to a variable in R

Single_t # we can call the variable for the results of the t test
```
Our t test is significant! p < 0.05, with a large t value. 


Using the {report} package will tell you if your t-test is significant and write up the results for you!
```{r}
library(report)
#report(t.test(df$depression, mu = 25)) - you can type in the thing, or just Single_t
report(Single_t)
```

You can write it up as this ^^^^ or you can also write it like this:

The mean depression score of college students (M = 41.07) was significantly greater than the mean depression score reported for the general population (M = 25), t(440) = 27.72, p > .001, Cohen's d = 1.32, 95 CI = 1.19, 1.45).

The {ggstatplot} will run the t test for you, show you a histogram of 
your distribution, and show you the mean of your sample, and normal distribution.
```{r}
library('ggstatsplot')
gghistostats(
  data       = df, # Put your data here
  x          = depression,  # put your x variable that you are running the test here
  title      = "Scores of Depression in College Age Students", # change the title
  test.value = 25, # This is the test statistic
  #Don't worry about the stuff below!
  binwidth   = 1,
  effsize.type = "d", # Effect size (i.e., how significant it is)
  normal.curve = TRUE, # adding a normal curve to it
  normal.curve.args = list(color = "darkred",linewidth = 1),
  bf.message = FALSE,
)
```

## Independent t-test 
Using base R, lets see if our depression scores differ between two groups (Gender)

1. We must first see if the varaince is equal or not. 
```{r}
# Template - var.test(testvar ~ groupvar, data = datasetname)
var.test(depression ~ male, data = df)
```

Equal variance is p>.05 so we can assume variance!

2. Doing the T test
```{r}
#template - t.test(testvar ~ groupvar, data = datasetname, var.equal = TRUE or FALSE)
t.test(depression ~ male, data = df, var.equal = TRUE)
```

3. Reporting
```{r}
library(report)
report(t.test(depression ~ male, data = df, var.equal = TRUE))
```

You can write it up as this with some minor changes or you can write it up like this:

An independent samples t-test was conducted to determine if the difference between the means on depression for males and females was significant. In comparison to males (M = 37.6), females (M = 42.2) report significantly higher depression symptomatology, t(439) = 3.438, p < .001; Cohen's d = 0.33, 95% CI = 0.14, 0.52).


4. vizualizing difference between groups
```{r}
library(ggstatsplot)
ggbetweenstats(
  data  = df,
  x     = male,
  y     = depression,
  title = "Distribution of Depression based on Gender",
  var.equal = TRUE,
  # You don't have to change below!
    bf.message = FALSE, # takes away stats you don't need
      effsize.type = "d" # Effect size (i.e., how significant it is)
)
```


## paired t-test
Say we wanted to look at people with depression before and after treatment.

1. Run the test. No variance needed.
```{r}
t.test(df$depression, df$depression2, paired = TRUE)
```

2. Report the finding
```{r}
report(t.test(df$depression, df$depression2, paired = TRUE))
```

You can also write it up as this:

A paired t-test was conducted to evaluate differences in depression scores over the course of treatment. The pre depression scores (M = 41.1, SD = 12.1) was significantly higher than post depression scores (M = 26.2, SD = 13.4, t(440) = 63.7, p  < 0.01; Cohen's d = 3.03, 95% CI = 2.81, 3.25). 

3. Visualize the finding

I couldn't get ggstatplot to work for paired t test :( I will find something and provide it for y'all if you are interested.




