---
title: "Practice quiz"
output: html_document
date: "2024-09-21"
editor_options: 
  markdown: 
    wrap: 72
---

# Practice Quiz 1 - Data Analysis Questions

These are the questions and answers for the first practice exam.

## Setup

### Install packages

```{r setup, include=FALSE}
if(!require("haven")) install.packages("haven")
if(!require('dplyr')) install.packages('dplyr')
if(!require('tidyverse')) {install.packages("tidyverse")}
if(!require('report')) {install.packages("report")}
if(!require('rstatix')) {install.packages("rstatix")}
if(!require('coin')) {install.packages("coin")}
```

### Load Packages

```{r}
library('haven') # import SPSS files
library('dplyr') #use this for data management
library('tidyverse') # for data stuff
library('report') # for fancy t test reporting
library('rstatix') # for wilcoxan stuff
library('ggpubr') # for plots
library('coin')
```

### Import data

```{r}
path <- file.choose()
path
df <- read_sav(path)
```

### Prep data for anlyses

```{r}
df$hadcovid = as.factor(df$hadcovid)
df$group = as.factor(df$group)
df
```

## Practice Exam Questions

1.  Run a one-sample t-test using the DV “time” and a test value of 50.
    Choose the correct result from the test below.

    a.  t(2389) = -2.600, p = .082
    b.  t(120) = 18.872, p \< .001
    c.  t(440) = -2.600, p = 0.010 \<- correct answer
    d.  t(440) = 2.620, p = 0.869

```{r}
q1 <- t.test(df$time, mu = 50)
q1
report(q1)
```

2.  Run an independent samples t-test using the IV “hadcovid” and the DV
    “time”. Choose the correct result from the test below.

    a.  t(439) = -1.567, p = .118 \<- correct answerr
    b.  t(232.760) = 5.692, p = .012
    c.  t(123.450) = 2.001, p = .052
    d.  t(440) = -1.635, p = .086

```{r}
var.test(time ~ hadcovid, data = df)
#not sig = vars are equal 
q2 <- t.test(time ~ hadcovid, data = df, var.equal = TRUE)
q2
report(q2)
```

3.  Run a dependent samples t-test using the DV “beforeanxiety” and the
    DV “afteranxiety”. Choose the correct result from the test below.
    Put in “beforeanxiety” first so the t value sign is correct.

    a.  t(440) = 25.325, p \< .001 \<- Correct answer
    b.  t(440) = 6.622, p = .912
    c.  t(167) = 0.561, p = .772
    d.  t(167) = 2.635, p = .046

```{r}
t.test(df$beforeanxiety, df$afteranxiety, paired = TRUE)
report(t.test(df$beforeanxiety, df$afteranxiety, paired = TRUE))
```

4.  Run a 3+ group Chi-Square test using the variable “group”. The
    hypothesized distribution for this variable is 33/33/33 or ‘equal’.
    Choose the correct result from the test below.

    a\. Chi-square value = 0.054, sig = .973 \<- Correct

    b\. Chi-square value = 1.504, sig = .773

    c\. Chi-square value = 2.789, sig = .005

    d\. Chi-square value = 0.632, sig \< .001

```{r}
table(df$group)
group<-c(147,145,149)
#run the chi square with the 57% female and 43% male expectation
q4  <- chisq.test(group, p = c((1/3), (1/3), (1/3)))
q4

#use report to tell us what this means!
report(q4)
```

5.  Run a Binomial Sign Test using the variable “covidsafety” and a test
    value of 15. Choose the correct result from the test below.

    a\. Test Statistic = 17.000, sig \< .001 \<- Correct

    b\. Test Statistic = 2.464, sig = 0.563

    c\. Test Statistic = 22.669, sig = 0.957

    d\. Test Statistic = 0.032, sig = .049

```{r}
df$success<- ifelse(df$covidsafety > 15,"above","below") 
#table
table(df2$success)
#length
length <- length(df2$success)
names(length) = "n"
length
#test
binom.test(x = 27,n =29, p =.5)
```

6.  Run a Mood’s Median Test using the IV “hadcovid” and the DV
    “covidsafety”. Choose the correct result from the test below.

    a\. Significance = .002 \<- Correct

    b\. Significance = .982

    c\. Significance = .125

    d\. Significance = .563

```{r}
median_test(df$covidsafety ~ df$hadcovid)
```

7.  Run a Mann Whitney U Test using the IV “hadcovid” and the DV
    “covidsafety”. Choose the correct result from the test below.

    a\. Z = -3.83, sig \< .001 \<- Correct

    b\. Z = -2.22, sig = .06

    c\. Z = -1.2, sig = .46

    d\. Z = -.5, sig = .72

```{r}
# run stats
wilcox.test(df$covidsafety ~ df$hadcovid)
# z test
z = as.numeric(statistic(wilcox_test(df2$depression ~ df2$male), type="standardized"))
names(z) = "z"
z
```
