---
title: "Practice quiz"
output: html_document
date: "2024-09-21"
editor_options: 
  markdown: 
    wrap: 72
---

# Practice Quiz 1 - Data Analysis Questions

These are the questions and answers for the first practice exam.

## Setup

### Install packages

```{r setup, include=FALSE}
if(!require("haven")) install.packages("haven")
if(!require('dplyr')) install.packages('dplyr')
if(!require('tidyverse')) {install.packages("tidyverse")}
if(!require('report')) {install.packages("report")}
if(!require(rstatix)) {install.packages("rstatix")}
if(!require(coin)) {install.packages("coin")}

```

### Load Packages 

```{r}
library('haven') # import SPSS files
library('dplyr') #use this for data management
library('tidyverse') # for data stuff
library('report') # for fancy t test reporting
library('rstatix') # for wilcoxan stuff
library('ggpubr') # for plots
library('coin')
```

### Import data

```{r}
path <- file.choose()
df <- read_sav(path)
```

### Prep data for anlyses

```{r}
df$hadcovid = as.factor(df$hadcovid)
df$group = as.factor(df$group)
df
```

## Practice Exam Questions

1.  Run a one-sample t-test using the DV “time” and a test value of 50.
    Choose the correct result from the test below.
    1.  t(2389) = -2.600, p = .082
    2.  t(120) = 18.872, p \< .001
    3.  t(440) = -2.600, p = 0.010 \<- correct answer
    4.  t(440) = 2.620, p = 0.869

```{r}
q1 <- t.test(df$time, mu = 50)
q1
report(q1)
```

2.  Run a dependent samples t-test using the DV “beforeanxiety” and the
    DV “afteranxiety”. Choose the correct result from the test below.
    Put in “beforeanxiety” first so the t value sign is correct.

    a\. t(440) = 25.325, p \< .001 \<- Correct answer

    b\. t(440) = 6.622, p = .912

    c\. t(167) = 0.561, p = .772

    d\. t(167) = 2.635, p = .046

```{r}
t.test(df$beforeanxiety, df$afteranxiety, paired = TRUE)
report(t.test(df$beforeanxiety, df$afteranxiety, paired = TRUE))

```

3.  Run a 3+ group Chi-Square test using the variable “group”. The
    hypothesized distribution for this variable is 33/33/33 or ‘equal’.
    Choose the correct result from the test below.

    m\. Chi-square value = 0.054, sig = .973 \<- Correct

    n\. Chi-square value = 1.504, sig = .773

    o\. Chi-square value = 2.789, sig = .005

    p\. Chi-square value = 0.632, sig \< .001

```{r}

```
